Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-03 paper/paper.tex
slight tweaks to abstract (moved immunopathology to first paragraph that talks about general consequences of IFN)

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some infected cells trigger innate immunity and begin producing interferon~\citep[IFN;][]{iwasaki2014innate}.
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}.

Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It remains unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
Such self-imposed limitation of an IFN response is consistent with the biological impact of IFN as, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology~\citep{iwasaki2014innate}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 virus-infected cells, 49 of which express IFN.
Even though we used a nominally ``wildtype'' stock of virus, two-thirds of the cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
Influenza infection only rarely triggers expression of the anti-viral IFNs~\citep[type I and type III IFNs;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
Importantly, the paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

>>>>

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some infected cells trigger innate immunity and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive prolonged immune responses later in infection can actually be associated with immunopathology and more severe disease~\citep{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a nominally ``wildtype'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
